---
title: "Practicum2"
author: "Thomas Faria, Caitlin Kirkpatrick, Sairah Shir"
date: "2024-07-14"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required packages for Practicum2
library(tidyverse)
library(readr)
library(RSQLite)

```

## Data Loading & Exploration

```{r, load}

# Load data
bed_type_df <- read_csv("data/bed_type.csv", show_col_types = FALSE)
bed_fact_df <- read_csv("data/bed_fact.csv", show_col_types = FALSE)
business_df <- read_csv("data/business.csv", show_col_types = FALSE)

```

```{r, explore}

# Explore data

# Summary of bed_type_df
summary(bed_type_df)
# Summary of bed_fact_df
summary(bed_fact_df)
# Summary of business_df
summary(business_df)

# Histograms for outlier visualization
# License
ggplot(bed_fact_df, aes(x = license_beds)) +
  geom_histogram(binwidth = 2, color = "black") +
  scale_x_continuous(limits = c(-1, 2100)) +
  scale_y_continuous(limits = c(-1, 3000)) +
  labs(title = "Data Distribution - bed_fact.csv - License", 
       x = "# of License Beds", 
       y = "Frequency",
       caption = "Figure 1. Histogram showing distrubtion of license bed counts") +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", 
                              margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", 
                              margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", 
                            margin = margin(b = 20)),
  plot.margin = margin(t = 30, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  )

# Census
ggplot(bed_fact_df, aes(x = census_beds)) +
  geom_histogram(binwidth = 2, color = "black") +
  scale_x_continuous(limits = c(-1, 2100)) +
  scale_y_continuous(limits = c(-1, 3000)) +
  labs(title = "Data Distribution - bed_fact.csv - License", 
       x = "# of License Beds", 
       y = "Frequency",
       caption = "Figure 2. Histogram showing distrubtion of license bed counts") +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", 
                              margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", 
                              margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", 
                            margin = margin(b = 20)),
  plot.margin = margin(t = 30, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  )

# Staffed
ggplot(bed_fact_df, aes(x = staffed_beds)) +
  geom_histogram(binwidth = 2, color = "black") +
  scale_x_continuous(limits = c(-1, 2100)) +
  scale_y_continuous(limits = c(-1, 3000)) +
  labs(title = "Data Distribution - bed_fact.csv - License", 
       x = "# of License Beds", 
       y = "Frequency",
       caption = "Figure 3. Histogram showing distrubtion of license bed counts") +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", 
                              margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", 
                              margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", 
                            margin = margin(b = 20)),
  plot.margin = margin(t = 30, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  )

```

### Create SQL Database

-   bed_fact_table, bed_type_table, and business_table are loaded from their corresponding R data frames into the "p2_db" database and a connection is established for later reference

```{r, create_db}

# Add data to a new database for downstream SQL queries
# Establish and open connection to existing database file
p2_connection <- dbConnect(RSQLite::SQLite(), "p2_db.sqlite")

# Function to check if table already exists in p2_db
table_exists <- function(con, table_name) {
  query <- sprintf("SELECT name FROM sqlite_master WHERE type='table' AND name='%s';", table_name)
  result <- dbGetQuery(con, query)
  return(nrow(result) > 0)
}

# Create bed_type_table, if it doesn't already exist
if (!table_exists(p2_connection, "bed_type_table")) {
  dbExecute(p2_connection, "
    CREATE TABLE bed_type_table (
      bed_id INTEGER PRIMARY KEY,
      bed_code TEXT,
      bed_desc TEXT
    );
  ")
}

# Create bed_fact_table, if it doesn't already exist
if (!table_exists(p2_connection, "bed_fact_table")) {
  dbExecute(p2_connection, "
    CREATE TABLE bed_fact_table (
      ims_org_id TEXT,
      bed_id INTEGER,
      license_beds INTEGER,
      census_beds INTEGER,
      staffed_beds INTEGER,
      PRIMARY KEY (ims_org_id, bed_id)
    );
  ")
}

# Create business_table, if it doesn't already exist
if (!table_exists(p2_connection, "business_table")) {
  dbExecute(p2_connection, "
    CREATE TABLE business_table (
      ims_org_id TEXT,
      business_name TEXT,
      ttl_license_beds INTEGER,
      ttl_census_beds INTEGER,
      ttl_staffed_beds INTEGER,
      bed_cluster_id INTEGER,
      PRIMARY KEY (ims_org_id, bed_cluster_id)
    );
  ")
}

# Load data into each table
dbWriteTable(p2_connection, "bed_type_table", bed_type_df, overwrite = TRUE, row.names = FALSE)
dbWriteTable(p2_connection, "bed_fact_table", bed_fact_df, overwrite = TRUE, row.names = FALSE)
dbWriteTable(p2_connection, "business_table", business_df, overwrite = TRUE, row.names = FALSE)

```

### Using JOIN to Create a Combined Table

-   combined_table is created from the existing fact/dimension tables and then loaded into combined_df for downstream R analysis

```{r, combined_table}

# Create combined_table, if it doesn't already exist
if (!table_exists(p2_connection, "combined_table")) {
  dbExecute(p2_connection, "
    CREATE TABLE combined_table (
      ims_org_id TEXT,
      business_name TEXT,
      ttl_license_beds INTEGER,
      ttl_census_beds INTEGER,
      ttl_staffed_beds INTEGER,
      bed_cluster_id INTEGER,
      bed_id INTEGER,
      license_beds INTEGER,
      census_beds INTEGER,
      staffed_beds INTEGER,
      bed_code TEXT,
      bed_desc TEXT,
      PRIMARY KEY (ims_org_id, bed_id)
    );
  ")
  # Insert data from other db tables
  dbExecute(p2_connection, "
    INSERT INTO combined_table
    SELECT business_table.ims_org_id,
           business_table.business_name,
           business_table.ttl_license_beds,
           business_table.ttl_census_beds,
           business_table.ttl_staffed_beds,
           business_table.bed_cluster_id,
           bed_fact_table.bed_id,
           bed_fact_table.license_beds,
           bed_fact_table.census_beds,
           bed_fact_table.staffed_beds,
           bed_type_table.bed_code,
           bed_type_table.bed_desc
    FROM business_table
    JOIN bed_fact_table ON business_table.ims_org_id = bed_fact_table.ims_org_id
    JOIN bed_type_table ON bed_fact_table.bed_id = bed_type_table.bed_id;
  ")
}

# Create combined dataframe
combined_df <- dbGetQuery(p2_connection, "SELECT * FROM combined_table")
head(combined_df)

```

### Outlier Identification

```{r, outlier_id}

# Compute outliers for each bed type in the combined_df
bed_type_outliers <- combined_df %>%
  mutate(
    mean_license_beds = mean(ttl_license_beds, na.rm = TRUE),
    mean_census_beds = mean(ttl_census_beds, na.rm = TRUE),
    mean_staffed_beds = mean(ttl_staffed_beds, na.rm = TRUE),
    sd_license_beds = sd(ttl_license_beds, na.rm = TRUE),
    sd_census_beds = sd(ttl_census_beds, na.rm = TRUE),
    sd_staffed_beds = sd(ttl_staffed_beds, na.rm = TRUE)
  ) %>%
  # Relative to the mean, any values on the lower or upper bounds that are 3 times the standard deviation are filtered
  filter(ttl_license_beds < mean_license_beds - 3 * sd_license_beds | ttl_license_beds > mean_license_beds + 3 * sd_license_beds) %>%
  filter(ttl_census_beds < mean_census_beds - 3 * sd_census_beds | ttl_census_beds > mean_census_beds + 3 * sd_census_beds) %>%
  filter(ttl_staffed_beds < mean_staffed_beds - 3 * sd_staffed_beds | ttl_staffed_beds > mean_staffed_beds + 3 * sd_staffed_beds) %>%
  select(ttl_license_beds, ttl_census_beds, ttl_staffed_beds)

summary(bed_type_outliers)

```

-   Outliers exist for staffed, license, and census bed counts within the data; however, we will not be excluding outliers from this analysis as any bed count should be included in this analysis to provide an optimal recommendation for the intervention program

### Verify Composite PKs

```{r, check_composite_pks}

# Verify the uniqueness of the composite primary keys used in bed_fact_table (combination of ims_org_id and bed_id) and business_table (combination of ims_org_id and bed_cluster_id)

duplicate_keys <- dbGetQuery(p2_connection, "
                             SELECT ims_org_id, bed_id, COUNT(*)
                             FROM bed_fact_table
                             GROUP BY ims_org_id, bed_id
                             HAVING COUNT(*) > 1;
                             ")

if (nrow(duplicate_keys) == 0) {
  print("Composite PK is unique for all combinations of ims_org_id and bed_id in bed_fact_table.")
} else {
  print("Duplicate keys found in bed_fact_table !! Composite PK is not valid.")
  print(duplicate_keys)
}

duplicate_keys <- dbGetQuery(p2_connection, "
                             SELECT ims_org_id, bed_cluster_id, COUNT(*)
                             FROM business_table
                             GROUP BY ims_org_id, bed_cluster_id
                             HAVING COUNT(*) > 1;
                             ")

if (nrow(duplicate_keys) == 0) {
  print("Composite PK is unique for all combinations of ims_org_id and bed_cluster_id in business_table.")
} else {
  print("Duplicate keys found in business_table !! Composite PK is not valid.")
  print(duplicate_keys)
}

```

## Identification of Facts/Dimensions in bed_type.csv and business.csv

### Dimensions

-   bed_type.csv contains three dimensions that are qualitative attributes providing context for or categorizing a fact object:
    -   bed_id (primary key) assigns a unique identifier for each type of bed
    -   bed_code contains abbreviations for each type of bed
    -   bed_desc contains descriptions of each type of bed
-   business.csv contains three dimensions that are qualitative attributes providing context for or categorizing a fact object:
    -   ims_org_id (primary key) assigns a unique identifier for each business
    -   business_name contains the names of the organizations
    -   bed_cluster_id categorizes bed clusters from 1-9

### Facts

-   The remaining columns in business.csv (ttl_license_beds, ttl_census_beds, and ttl_staffed_bed)s are facts due to their quantitative nature

## Identification of Facts/Dimensions in bed_fact.csv

### Dimensions

-   bed_fact.csv contains one dimension that is qualitative:
    -   ims_org_id (primary key) is a dimension as it assigns a unique identifier for each business

### Facts

-   bed_fact.csv contains four facts:
    -   bed_id, license_beds, census_beds, staffed_beds are all facts since they report quantitative data

------------------------------------------------------------------------

## Initial Analysis for Leadership

### Example Query for Section 3a:

``` sql
SELECT business_name AS hospital_name,
--- Take the sum of the staffed beds in a new column total_icu_sicu_staffed_beds ---
SUM(staffed_beds) AS total_icu_sicu_staffed_beds
FROM combined_table
--- Only include rows where bed_id is 4 (ICU) or 15 (SICU) ---
WHERE bed_id IN (4, 15)
--- Group by hospital_name ---
GROUP BY hospital_name
--- Arrange in descending order ---
ORDER BY total_icu_sicu_staffed_beds DESC
--- Only include top 10 hospitals ---
LIMIT 10;
```

### License Beds (ICU or SICU)

```{r, license_beds_total}

get_license_beds_total <- "SELECT business_name AS hospital_name,
SUM(staffed_beds) AS total_icu_sicu_license_beds
FROM combined_table
WHERE bed_id IN (4, 15)
GROUP BY hospital_name
ORDER BY total_icu_sicu_license_beds DESC
LIMIT 10;"

license_beds_total_df <- dbGetQuery(p2_connection, get_license_beds_total)
license_beds_total_df

```

```{r, license_beds_visual}

separate_license_beds_visual <- "SELECT business_name AS hospital_name,
--- When bed_desc is ICU, the value is added to the total, else 0 is added ---
SUM(CASE WHEN bed_desc = 'ICU' THEN license_beds ELSE 0 END) AS ICU_beds,
--- When bed_desc is SICU, the value is added to the total, else 0 is added ---
SUM(CASE WHEN bed_desc = 'SICU' THEN license_beds ELSE 0 END) AS SICU_beds,
SUM(census_beds) AS total_icu_sicu_license_beds
FROM combined_table
--- Only include rows where bed_id is 4 (ICU) or 15 (SICU) ---
WHERE bed_desc IN ('ICU', 'SICU')
GROUP BY hospital_name
--- Arrange in descending order ---
ORDER BY total_icu_sicu_license_beds DESC;"

# Create new dataframe for visual
separate_license_beds_df <- dbGetQuery(p2_connection, separate_license_beds_visual) %>%
  as_tibble() %>%
  slice(1:10) %>%
  rename(ICU = ICU_beds,
         SICU = SICU_beds)
separate_license_beds_df

pivot_license_beds_df <- separate_license_beds_df %>%
  pivot_longer(cols = c(ICU, SICU), names_to = "bed_type", values_to = "bed_count")

```

```{r, license_beds_visual_ggplot}

# Stacked bar for better visualization of ICU and SICU beds for top 10 hospitals
ggplot(pivot_license_beds_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = bed_type)) +
  geom_bar(stat = "identity") +
  labs(
  x = "Hospital Name",
  y = "License Bed Count",
  fill = "Bed Type",
  title = "Hospitals with Highest Total ICU or SICU License Beds",
  caption = "Figure 4: ICU and SICU license beds combined with no minimum required count"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", margin = margin(b = 10)),
  plot.margin = margin(t = 20, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  ) +
  coord_flip()
```

### Census Beds (ICU or SICU)

```{r, census_beds_total}

get_census_beds_total <- "SELECT business_name AS hospital_name, 
SUM(census_beds) AS total_icu_sicu_census_beds
FROM combined_table
WHERE bed_id IN (4, 15)
GROUP BY hospital_name
ORDER BY total_icu_sicu_census_beds DESC
LIMIT 10"

census_beds_total_df <- dbGetQuery(p2_connection, get_census_beds_total)
census_beds_total_df

```

```{r, census_beds_visual}

separate_census_beds_visual <- "SELECT business_name AS hospital_name,
--- When bed_desc is ICU, the value is added to the total, else 0 is added ---
SUM(CASE WHEN bed_desc = 'ICU' THEN census_beds ELSE 0 END) AS ICU_beds,
--- When bed_desc is SICU, the value is added to the total, else 0 is added ---
SUM(CASE WHEN bed_desc = 'SICU' THEN census_beds ELSE 0 END) AS SICU_beds,
SUM(census_beds) AS total_icu_sicu_census_beds
FROM combined_table
--- Only include rows where bed_id is 4 (ICU) or 15 (SICU) ---
WHERE bed_desc IN ('ICU', 'SICU')
GROUP BY hospital_name
--- Arrange in descending order ---
ORDER BY total_icu_sicu_census_beds DESC;"

# Create new dataframe for visual
separate_census_beds_df <- dbGetQuery(p2_connection, separate_census_beds_visual) %>%
  as_tibble() %>%
  slice(1:10) %>%
  rename(ICU = ICU_beds,
         SICU = SICU_beds)
separate_census_beds_df

pivot_census_beds_df <- separate_census_beds_df %>%
  pivot_longer(cols = c(ICU, SICU), names_to = "bed_type", values_to = "bed_count")

```

```{r, census_visual_ggplot}

# Stacked bar for better visualization of ICU and SICU beds for top 10 hospitals
ggplot(pivot_census_beds_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = bed_type)) +
  geom_bar(stat = "identity") +
  labs(
  x = "Hospital Name",
  y = "Census Bed Count",
  fill = "Bed Type",
  title = "Hospitals with Highest Total ICU or SICU Census Beds",
  caption = "Figure 5: ICU and SICU census beds combined with no minimum required count"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", margin = margin(b = 10)),
  plot.margin = margin(t = 20, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  ) +
  coord_flip()
```

### Staffed Beds (ICU or SICU)

```{r, staffed_beds_total}

get_staffed_beds_total <- "SELECT business_name AS hospital_name,
--- Take the sum of the staffed beds in a new column total_icu_sicu_staffed_beds ---
SUM(staffed_beds) AS total_icu_sicu_staffed_beds
FROM combined_table
--- Only include rows where bed_id is 4 (ICU) or 15 (SICU) ---
WHERE bed_id IN (4, 15)
--- Group by hospital_name ---
GROUP BY hospital_name
--- Arrange in descending order ---
ORDER BY total_icu_sicu_staffed_beds DESC
--- Only include top 10 hospitals ---
LIMIT 10;"

total_staffed_beds <- dbGetQuery(p2_connection, get_staffed_beds_total) %>%
  as_tibble() %>%
  mutate(
    total_icu_sicu_staffed_beds = as.numeric(total_icu_sicu_staffed_beds)
  )
total_staffed_beds
```

```{r staffed_visual}

# Query for the staffed visual
staffed_beds_separate_visual <- "SELECT business_name AS hospital_name,
                                 --- When bed_desc is ICU, the value is added to the total, else 0 is added ---
                                 SUM(CASE WHEN bed_desc = 'ICU' THEN staffed_beds ELSE 0 END) AS ICU_beds,
                                 --- When bed_desc is SICU, the value is added to the total, else 0 is added ---
                                 SUM(CASE WHEN bed_desc = 'SICU' THEN staffed_beds ELSE 0 END) AS SICU_beds,
                                 SUM(staffed_beds) AS total_icu_sicu_staffed_beds
                                 FROM combined_table
                                 --- Only include rows where bed_id is 4 (ICU) or 15 (SICU) ---
                                 WHERE bed_desc IN ('ICU', 'SICU')
                                 GROUP BY hospital_name
                                 --- Arrange in descending order ---
                                 ORDER BY total_icu_sicu_staffed_beds DESC;"

# Create new dataframe for visual
separate_staffed_beds_df <- dbGetQuery(p2_connection, staffed_beds_separate_visual) %>%
  as_tibble() %>%
  slice(1:10) %>%
  rename(ICU = ICU_beds,
         SICU = SICU_beds)
separate_staffed_beds_df

pivot_staffed_beds_df <- separate_staffed_beds_df %>%
  pivot_longer(cols = c(ICU, SICU), names_to = "bed_type", values_to = "bed_count")
```

```{r, staffed_visual_ggplot}

# Stacked bar for better visualization of ICU and SICU beds for top 10 hospitals
ggplot(pivot_staffed_beds_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = bed_type)) +
  geom_bar(stat = "identity") +
  labs(
  x = "Hospital Name",
  y = "Staffed Bed Count",
  fill = "Bed",
  title = "Hospitals with highest combined ICU or SICU staffed bed counts",
  caption = "Figure 6: ICU and SICU staffed beds combined with no minimum required count"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", margin = margin(b = 10)),
  plot.margin = margin(t = 20, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  ) +
  coord_flip()
```

### Interpretation of Findings (Pt. 1)

The hospitals that have the most license ICU and SICU beds are Phoenix Childrens Hospital and University of Maryland Medical Center. Based on the finding for the census beds, it appears Shands Hospital at University of Florida and Dallas County Hospital Association have the most census beds that include ICU and SICU beds.

The hospitals that have the most staffed beds are Vidant Medical Center and Rady Childrens Hospital and Health Center. Some hospitals show up on all three tables in the top 10 hospitals for each of the different beds. For example, Vidant Medical Center appears on all three tables and could be a good candidate for the intervention pilot program.

It should be noted that despite having the highest number of ICU and SICU beds amongst the staffed, census, and license beds, Vidant Medical Center and Rady Childrens Hospital and Health Center do not appear to have an even distribution of both ICU and SICU beds.

------------------------------------------------------------------------

## Follow-up Analysis for Leadership (identifying hospitals with at least 1 of both ICU and SICU)

### License Beds (ICU and SICU Combined - min. 1 bed of each type per hospital)

```{r, license_beds_min1}
get_license_beds_min1 <- "SELECT business_name AS hospital_name, 
SUM(license_beds) AS total_license_beds_min1
FROM combined_table
WHERE (bed_id = 4 OR bed_id = 15)
GROUP BY hospital_name
HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
ORDER BY total_license_beds_min1 DESC
LIMIT 10;"

license_beds_min1_df <- dbGetQuery(p2_connection, get_license_beds_min1) %>%
  as_tibble()
license_beds_min1_df

```

```{r, license_beds_min1_visual}

license_beds_min1_visual <- "SELECT business_name AS hospital_name,
                                 --- When bed_id is 4, the value is added to the total, else 0 is added ---
                                 SUM(CASE WHEN bed_id = 4 THEN license_beds ELSE 0 END) AS ICU_beds,
                                 --- When bed_id is 15, the value is added to the total, else 0 is added ---
                                 SUM(CASE WHEN bed_id = 15 THEN license_beds ELSE 0 END) AS SICU_beds,
                                 SUM(license_beds) AS total_icu_sicu_beds
                             FROM combined_table
                             --- Include ICU and SICU ---
                             WHERE (bed_id = 4 OR bed_id = 15)
                             GROUP BY hospital_name
                             --- Ensures at least 1 bed is counted in SICU and ICU ---
                             HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
                               AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
                             ORDER BY total_icu_sicu_beds DESC;"

license_beds_min1_visual_df <- dbGetQuery(p2_connection, license_beds_min1_visual) %>%
  as_tibble %>%
  slice(1:10) %>%
  rename(ICU = ICU_beds,
         SICU = SICU_beds)

license_beds_min1_visual_df

pivot_min1_license_beds_visual_df <- license_beds_min1_visual_df %>%
  pivot_longer(cols = c(ICU, SICU), names_to = "bed_type", values_to = "bed_count")
```

```{r, license_beds_min1_visual_ggplot}

# Stacked bar for better visualization of ICU and SICU beds for top 10 hospitals
ggplot(pivot_min1_license_beds_visual_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = bed_type)) +
  geom_bar(stat = "identity") +
  labs(
  x = "Hospital Name",
  y = "license Bed Count",
  fill = "Bed Type",
  title = "Hospitals with Highest Total ICU and SICU License Beds (Min. 1)",
  caption = "Figure 7: Minimum 1 ICU and SICU license bed per hospital"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", margin = margin(b = 10)),
  plot.margin = margin(t = 20, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  ) +
  coord_flip()

```

### Census Beds (ICU and SICU Combined - min. 1 bed of each type per hospital)

```{r, census_beds_min1}

census_beds_min1 <- "SELECT business_name AS hospital_name, 
SUM(census_beds) AS total_census_beds_min1
FROM combined_table
WHERE (bed_id = 4 OR bed_id = 15)
GROUP BY hospital_name
HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
ORDER BY total_census_beds_min1 DESC
LIMIT 10;"

census_beds_min1_df <- dbGetQuery(p2_connection, census_beds_min1)
census_beds_min1_df
```

```{r, census_beds_min1_visual}

census_beds_min1_visual <- "SELECT business_name AS hospital_name,
--- When bed_id is 4, the value is added to the total, else 0 is added ---
SUM(CASE WHEN bed_id = 4 THEN census_beds ELSE 0 END) AS ICU_beds,
--- When bed_id is 15, the value is added to the total, else 0 is added ---
SUM(CASE WHEN bed_id = 15 THEN census_beds ELSE 0 END) AS SICU_beds,
SUM(census_beds) AS total_icu_sicu_beds
FROM combined_table
WHERE (bed_id = 4 OR bed_id = 15)
GROUP BY hospital_name
HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
ORDER BY total_icu_sicu_beds DESC
LIMIT 10;"

census_beds_min1_visual_df <- dbGetQuery(p2_connection, census_beds_min1_visual) %>%
  as_tibble %>%
  slice(1:10) %>%
  rename(ICU = ICU_beds,
         SICU = SICU_beds)

census_beds_min1_visual_df

pivot_min1_census_beds_visual_df <- census_beds_min1_visual_df %>%
  pivot_longer(cols = c(ICU, SICU), names_to = "bed_type", values_to = "bed_count")
```

```{r, census_beds_min1_visual_ggplot}

# Stacked bar for better visualization of ICU and SICU beds for top 10 hospitals
ggplot(pivot_min1_census_beds_visual_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = bed_type)) +
  geom_bar(stat = "identity") +
  labs(
  x = "Hospital Name",
  y = "Census Bed Count",
  fill = "Bed Type",
  title = "Hospitals with Highest Total ICU and SICU Census Beds (Min. 1)",
  caption = "Figure 8: Minimum 1 ICU and SICU census bed per hospital"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", margin = margin(b = 10)),
  plot.margin = margin(t = 20, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  ) +
  coord_flip()

```

### Staffed Beds (ICU and SICU Combined - min. 1 bed of each type per hospital)

```{r, staffed_beds_min1}

staffed_beds_min1 <- "SELECT business_name AS hospital_name, 
SUM(census_beds) AS total_staffed_beds_min1
FROM combined_table
WHERE (bed_id = 4 OR bed_id = 15)
GROUP BY hospital_name
HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
ORDER BY total_staffed_beds_min1 DESC
LIMIT 10;"

staffed_beds_min1_df <- dbGetQuery(p2_connection, staffed_beds_min1)
staffed_beds_min1_df
```

```{r staffed_beds_min1_visual}

staffed_beds_min1_visual <- "SELECT business_name AS hospital_name,
                                 --- When bed_id is 4, the value is added to the total, else 0 is added ---
                                 SUM(CASE WHEN bed_id = 4 THEN staffed_beds ELSE 0 END) AS ICU_beds,
                                 --- When bed_id is 15, the value is added to the total, else 0 is added ---
                                 SUM(CASE WHEN bed_id = 15 THEN staffed_beds ELSE 0 END) AS SICU_beds,
                                 SUM(staffed_beds) AS total_icu_sicu_beds
                          FROM combined_table
                          --- Include ICU and SICU ---
                          WHERE (bed_id = 4 OR bed_id = 15)
                          GROUP BY hospital_name
                          --- Ensures at least 1 bed is counted in SICU and ICU ---
                          HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
                             AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
                          ORDER BY total_icu_sicu_beds DESC;"

staffed_beds_min1_visual_df <- dbGetQuery(p2_connection, staffed_beds_min1_visual) %>%
  as_tibble %>%
  slice(1:10) %>%
  rename(ICU = ICU_beds,
         SICU = SICU_beds)

staffed_beds_min1_visual_df

pivot_min1_staffed_beds_visual_df <- staffed_beds_min1_visual_df %>%
  pivot_longer(cols = c(ICU, SICU), names_to = "bed_type", values_to = "bed_count")

```

```{r staffed_beds_min1_visual_ggplot}

# Stacked bar for better visualization of ICU and SICU beds for top 10 hospitals
ggplot(pivot_min1_staffed_beds_visual_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = bed_type)) +
  geom_bar(stat = "identity") +
  labs(
  x = "Hospital Name",
  y = "Staffed Bed Count",
  fill = "Bed",
  title = "Hospitals with Highest Total ICU and SICU Staffed Beds",
  caption = "Figure 9: Minimum 1 ICU and SICU staffed bed per hospital"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(size = 9, color = "black"),
  axis.title.x = element_text(color = "black", face = "bold", margin = margin(t = 5, b = 5)),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black", face = "bold", margin = margin(r = 10)),
  plot.title = element_text(hjust = 0.8, face = "bold", margin = margin(b = 10)),
  plot.margin = margin(t = 20, r = 20, l = 20, b = 10),
  plot.caption.position = "plot",
  plot.caption = element_text(face = "italic", hjust = 1)
  ) +
  coord_flip()

```

### Interpretation of Findings (Pt. 2)

Leadership was interested in hospitals that had a sufficient volume of both ICU and SICU beds and in doing this analysis for all three bed types we found a few recommendations.

-   License Beds

We maintain that the University of Maryland still has the highest combined total of license ICU and SICU beds. The share is evenly split between both bed types. While other hospitals may have comparable or more of one bed type vs. the other (UC Health University Hospital has more ICU, while Carolinas Medical Center has comparable SICU but far less ICU when evaluated side-by-side), Maryland remains the best-balanced choice.

-   Census Beds

We found from 3a that Shands Hospital at University of Florida had the most census ICU and SICU beds. However, with further investigation and making sure that the hospitals had at least one of each, ICU and SICU, we find that leadership might be interested in different hospitals. Shands Hospital at the University of Florida have more than one ICU and SICU census beds. However, there is not an equal number of ICU and SICU census beds. The best hospitals for this pilot intervention program would be Cedars Sinai Health System, Jackson Memorial Hospital, and University of Maryland Medical Center. These all have roughly the same amount of ICU and SICU census beds.

-   Staffed Beds

With the added condition that hospitals must have at least 1 ICU and SICU bed type, the resulting top 10 hospitals are quite different. 5 hospitals have been displaced that were previously shown in Figure 3C, the most notable being Vidant Medical Center and Ray Childrens Hospital and Health Center. Despite having a high count of ICU beds, neither had any SICU beds. In Figure 4C we can see the distribution of ICU and SICU staffed beds across the hospitals.

The University of Maryland Medical Center has an even distribution of ICU and SICU staffed beds. This is advantageous for cost-efficiency to ensure neither ICU or SICU beds counts are too low when rolling out the intervention program. Shands Hospital at the University of Florida has significantly more SICU beds than ICU beds, so despite having the second highest bed count, the need for a relatively even spread of ICU and SICU beds makes it difficult to recommend.

------------------------------------------------------------------------

## Final Recommendation

```{r combined_visual}

all_beds_combined_visual <- "SELECT business_name AS hospital_name,
                                 --- When bed_id is 4, the value is added to the total, else 0 is added (for each bed type, respectively) ---
                                 SUM(CASE WHEN bed_id = 4 THEN license_beds ELSE 0 END) AS ICU_license_beds,
                                 --- When bed_id is 15, the value is added to the total, else 0 is added (for each bed type, respectively) ---
                                 SUM(CASE WHEN bed_id = 15 THEN license_beds ELSE 0 END) AS SICU_license_beds,
                                 SUM(license_beds) AS total_license_icu_sicu_beds,
                                 SUM(CASE WHEN bed_id = 4 THEN census_beds ELSE 0 END) AS ICU_census_beds,
                                 SUM(CASE WHEN bed_id = 15 THEN census_beds ELSE 0 END) AS SICU_census_beds,
                                 SUM(census_beds) AS total_census_icu_sicu_beds,
                                 SUM(CASE WHEN bed_id = 4 THEN staffed_beds ELSE 0 END) AS ICU_staffed_beds,
                                 SUM(CASE WHEN bed_id = 15 THEN staffed_beds ELSE 0 END) AS SICU_staffed_beds,
                                 SUM(staffed_beds) AS total_staffed_icu_sicu_beds
                          FROM combined_table
                          --- Include ICU and SICU ---
                          WHERE (bed_id = 4 OR bed_id = 15)
                          GROUP BY hospital_name
                          --- Ensures at least 1 bed is counted in SICU and ICU ---
                          HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
                             AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
                          ORDER BY total_license_icu_sicu_beds + total_census_icu_sicu_beds + total_staffed_icu_sicu_beds DESC;"

all_beds_combined_visual_df <- dbGetQuery(p2_connection, all_beds_combined_visual) %>%
  as_tibble %>%
  slice(1:10) %>%
  rename("Census ICU" = ICU_census_beds,
         "Census SICU" = SICU_census_beds,
         "License ICU" = ICU_license_beds,
         "License SICU" = SICU_license_beds,
         "Staffed ICU" = ICU_staffed_beds,
         "Staffed SICU" = SICU_staffed_beds)

all_beds_combined_visual_df

pivot_all_beds_visual_df <- all_beds_combined_visual_df %>%
  pivot_longer(
    # Pivot License, Census, and Staffed ICU and SICU columns 
    cols = starts_with("License") | starts_with("Census") | starts_with("Staffed"),
    names_to = c("bed_type", "icu_sicu"),
    names_sep = " ",
    values_to = "bed_count"
  ) %>%
  # Create factor levels for stacked bar
  mutate(
    bed_type = factor(bed_type, levels = c("License", "Census", "Staffed")),
    icu_sicu = factor(icu_sicu, levels = c("ICU", "SICU"))
  )
```

```{r all_beds_visual}

ggplot(pivot_all_beds_visual_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = interaction(bed_type, icu_sicu, sep = " "))) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_wrap(~ icu_sicu, ncol = 1, scales = "fixed") +
  labs(
    x = "Hospital Name",
    y = "Bed Count",
    fill = "Bed Type",
    title = "Hospitals with Highest Combined Bed Type Count for ICU and SICU Beds",
    caption = "Figure 10: Bed type breakdown between ICU and SICU counts for top 10 hospitals"
  ) +
  theme_classic() +
  theme(
    axis.text = element_text(color = "black"),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    plot.title = element_text(hjust = 0.7, face = "bold", margin = margin(b = 10)),
    plot.margin = margin(t = 20, r = 20, l = 20, b = 10),
    plot.caption.position = "plot",
    plot.caption = element_text(face = "italic", hjust = 1)
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  coord_flip()
```

To maintain cost efficiency when implementing a hiring intervention we analyzed the volume of ICU and SICU beds across three bed types: license, census, and staffed beds. We derived the top 10 hospitals with sufficient volume for either SICU or ICU beds and the top 10 hospitals with at least 1 of both types.

We visualized the distribution of ICU and SICU bed counts with stacked bar charts for license, census, and staffed beds in Figure 10, respectively. We recommend considering hospitals that not only have at least 1 of each type of bed, but an equal distribution of both types to maximize cost-efficiency and avoid bed shortages. Because of this, we believe hospitals with high total bed counts are not necessarily better candidates if they only have more of one bed type. For example, the total bed count at Shands Hospital at The University of Florida in Figure 10 shows a majority SICU beds and small portion of ICU beds. We would not recommend this hospital due to the potential ICU bed shortages.

The two hospitals that best meet the cost-efficient criteria are The University of Maryland Medical Center and Jackson Memorial Hospital. We believe both offer a balanced distribution of ICU and SICU beds while maintaining an appropriate total bed count for beginning the intervention program.

------------------------------------------------------------------------

### Sources

1.  <https://medium.com/illumination/mastering-conditional-logic-in-sql-a-deep-dive-into-case-when-statements-52f0aca9890c>
2.  <https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/interaction>
3.  <https://ggplot2-book.org/facet>
