---
title: "Practicum2"
author: "Thomas Faria, Caitlin Kirkpatrick, Sairah Shir"
date: "2024-07-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required packages for Practicum2
library(tidyverse)
library(readr)
library(RSQLite)

```

### Loading and data exploration

```{r, load}

# Load data
bed_type_df <- read_csv("data/bed_type.csv", show_col_types = FALSE)
bed_fact_df <- read_csv("data/bed_fact.csv", show_col_types = FALSE)
business_df <- read_csv("data/business.csv", show_col_types = FALSE)

```

```{r, explore}

# Explore data
print("Summary of bed_type.csv")
summary(bed_type_df)
print("Summary of bed_fact.csv")
summary(bed_fact_df)
print("Summary of business.csv")
summary(business_df)

# Basic histograms for outlier visualization
ggplot(bed_fact_df, aes(x = license_beds)) +
  geom_histogram(binwidth = 2, color = "black") +
  scale_x_continuous(limits = c(-1, 3000)) +
  scale_y_continuous(limits = c(-1, 3000)) +
  labs(title = "Data Distribution - bed_fact.csv", x = "# of License Beds", y = "Frequency") +
  theme_minimal()

ggplot(bed_fact_df, aes(x = census_beds)) +
  geom_histogram(binwidth = 2, color = "black") +
  scale_x_continuous(limits = c(-1, 3000)) +
  scale_y_continuous(limits = c(-1, 3000)) +
  labs(title = "Data Distribution - bed_fact.csv", x = "# of Census Beds", y = "Frequency") +
  theme_minimal()

ggplot(bed_fact_df, aes(x = staffed_beds)) +
  geom_histogram(binwidth = 2, color = "black") +
  scale_x_continuous(limits = c(-1, 3000)) +
  scale_y_continuous(limits = c(-1, 3000)) +
  labs(title = "Data Distribution - bed_fact.csv", x = "# of Staffed Beds", y = "Frequency") +
  theme_minimal()

```

```{r, create_db}

# Add data to a new database for downstream SQL queries
# Establish and open connection to existing database file
p2_connection <- dbConnect(RSQLite::SQLite(), "p2_db.sqlite")

# Function to check if table already exists in p2_db
table_exists <- function(con, table_name) {
  query <- sprintf("SELECT name FROM sqlite_master WHERE type='table' AND name='%s';", table_name)
  result <- dbGetQuery(con, query)
  return(nrow(result) > 0)
}

# Create bed_type_table, if it doesn't already exist
if (!table_exists(p2_connection, "bed_type_table")) {
  dbExecute(p2_connection, "
    CREATE TABLE bed_type_table (
      bed_id INTEGER PRIMARY KEY,
      bed_code TEXT,
      bed_desc TEXT
    );
  ")
}

# Create bed_fact_table, if it doesn't already exist
if (!table_exists(p2_connection, "bed_fact_table")) {
  dbExecute(p2_connection, "
    CREATE TABLE bed_fact_table (
      ims_org_id TEXT,
      bed_id INTEGER,
      license_beds INTEGER,
      census_beds INTEGER,
      staffed_beds INTEGER,
      PRIMARY KEY (ims_org_id, bed_id)
    );
  ")
}

# Create business_table, if it doesn't already exist
if (!table_exists(p2_connection, "business_table")) {
  dbExecute(p2_connection, "
    CREATE TABLE business_table (
      ims_org_id TEXT,
      business_name TEXT,
      ttl_license_beds INTEGER,
      ttl_census_beds INTEGER,
      ttl_staffed_beds INTEGER,
      bed_cluster_id INTEGER,
      PRIMARY KEY (ims_org_id, bed_cluster_id)
    );
  ")
}

# Load data into each table
dbWriteTable(p2_connection, "bed_type_table", bed_type_df, overwrite = TRUE, row.names = FALSE)
dbWriteTable(p2_connection, "bed_fact_table", bed_fact_df, overwrite = TRUE, row.names = FALSE)
dbWriteTable(p2_connection, "business_table", business_df, overwrite = TRUE, row.names = FALSE)

```

* bed_fact_table, bed_type_table, and business_table are loaded from their corresponding R dataframes

```{r, combined_table}
# Create combined_table, if it doesn't already exist
if (!table_exists(p2_connection, "combined_table")) {
  dbExecute(p2_connection, "
    CREATE TABLE combined_table (
      ims_org_id TEXT,
      business_name TEXT,
      ttl_license_beds INTEGER,
      ttl_census_beds INTEGER,
      ttl_staffed_beds INTEGER,
      bed_cluster_id INTEGER,
      bed_id INTEGER,
      license_beds INTEGER,
      census_beds INTEGER,
      staffed_beds INTEGER,
      bed_code TEXT,
      bed_desc TEXT,
      PRIMARY KEY (ims_org_id, bed_id)
    );
  ")
  # Insert data from other db tables
  dbExecute(p2_connection, "
    INSERT INTO combined_table
    SELECT business_table.ims_org_id,
           business_table.business_name,
           business_table.ttl_license_beds,
           business_table.ttl_census_beds,
           business_table.ttl_staffed_beds,
           business_table.bed_cluster_id,
           bed_fact_table.bed_id,
           bed_fact_table.license_beds,
           bed_fact_table.census_beds,
           bed_fact_table.staffed_beds,
           bed_type_table.bed_code,
           bed_type_table.bed_desc
    FROM business_table
    JOIN bed_fact_table ON business_table.ims_org_id = bed_fact_table.ims_org_id
    JOIN bed_type_table ON bed_fact_table.bed_id = bed_type_table.bed_id;
  ")
}

# Create combined dataframe
combined_df <- dbGetQuery(p2_connection, "SELECT * FROM combined_table")
head(combined_df)
str(combined_df)

```

* combined_table is loaded from the existing db tables and then loaded into combined_df

```{r, tidy}

# Outlier identification
# Compute outliers for each bed type in combined_df
bed_type_outliers <- combined_df %>%
  mutate(
    mean_license_beds = mean(ttl_license_beds, na.rm = TRUE),
    mean_census_beds = mean(ttl_census_beds, na.rm = TRUE),
    mean_staffed_beds = mean(ttl_staffed_beds, na.rm = TRUE),
    sd_license_beds = sd(ttl_license_beds, na.rm = TRUE),
    sd_census_beds = sd(ttl_census_beds, na.rm = TRUE),
    sd_staffed_beds = sd(ttl_staffed_beds, na.rm = TRUE)
  ) %>%
  # Relative to the mean, any values on the lower or upper bounds that are 3 times the standard deviation are filtered
  filter(ttl_license_beds < mean_license_beds - 3 * sd_license_beds | ttl_license_beds > mean_license_beds + 3 * sd_license_beds) %>%
  filter(ttl_census_beds < mean_census_beds - 3 * sd_census_beds | ttl_census_beds > mean_census_beds + 3 * sd_census_beds) %>%
  filter(ttl_staffed_beds < mean_staffed_beds - 3 * sd_staffed_beds | ttl_staffed_beds > mean_staffed_beds + 3 * sd_staffed_beds) %>%
  select(ttl_license_beds, ttl_census_beds, ttl_staffed_beds)

bed_type_outliers

```

```{r, check_composite_pks}

# Verify the uniqueness of the composite primary keys used in bed_fact_table (combination of ims_org_id and bed_id) and business_table (combination of ims_org_id and bed_cluster_id)

duplicate_keys <- dbGetQuery(p2_connection, "
                             SELECT ims_org_id, bed_id, COUNT(*)
                             FROM bed_fact_table
                             GROUP BY ims_org_id, bed_id
                             HAVING COUNT(*) > 1;
                             ")

if (nrow(duplicate_keys) == 0) {
  print("Composite PK is unique for all combinations of ims_org_id and bed_id in bed_fact_table.")
} else {
  print("Duplicate keys found in bed_fact_table !! Composite PK is not valid.")
  print(duplicate_keys)
}

duplicate_keys <- dbGetQuery(p2_connection, "
                             SELECT ims_org_id, bed_cluster_id, COUNT(*)
                             FROM business_table
                             GROUP BY ims_org_id, bed_cluster_id
                             HAVING COUNT(*) > 1;
                             ")

if (nrow(duplicate_keys) == 0) {
  print("Composite PK is unique for all combinations of ims_org_id and bed_cluster_id in business_table.")
} else {
  print("Duplicate keys found in business_table !! Composite PK is not valid.")
  print(duplicate_keys)
}

```

### Identification of Dimensions in bed_type.csv and business.csv

```{r}
str(bed_type_df)
str(business_df)
```

-   bed_type_df contains three dimensions that are qualitative attributes providing context for or categorizing a fact object:
    -   bed_id (primary key) assigns a unique identifier for each type of bed
    -   bed_code contains abbreviations for each type of bed
    -   bed_desc contains descriptions of each type of bed
-   business_df contains three dimensions that are qualitative attributes providing context for or categorizing a fact object:
    -   ims_org_id (primary key) assigns a unique identifier for each business
    -   business_name contains the names of the organizations
    -   bed_cluster_id categorizes bed clusters from 1-9
-   The remaining columns ttl_license_beds, ttl_census_beds, and ttl_staffed_beds are facts due to their quantitative nature

### Identification of Fact Variables in bed_fact.csv

-   bed_fact_df has one dimension that is qualitative nature:
    -   ims_org_id (primary key) is a dimension as it assigns unique identifier for each business\
-   bed_fact_df has for facts:
    -   Bed_id, license_bed, census_bed, staffed_bed are all facts due to the numerical data

### Initial Analysis for Leadership

```{r, license_beds_separate}

license_beds_ICU <- "SELECT business_table.business_name AS hospital_name,
--- Take the sum of the license beds in a new column total_icu_license_beds ---
SUM(bed_fact_table.license_beds) AS total_icu_license_beds
FROM business_table
--- Inner join bed_fact_table with f on ims_org_id ---
JOIN bed_fact_table ON business_table.ims_org_id = bed_fact_table.ims_org_id
--- Only include rows where bed_id is 4 (ICU) ---
WHERE bed_fact_table.bed_id = 4
--- Group by hospital_name ---
GROUP BY business_table.business_name
--- Arrange in descending order ---
ORDER BY total_icu_license_beds DESC
--- Only include top 10 hospitals ---
LIMIT 10;"

license_beds_ICU_df <- dbGetQuery(p2_connection, license_beds_ICU)

license_beds_SICU <- "SELECT business_table.business_name AS hospital_name,
--- Take the sum of the license beds in a new column total_sicu_license_beds ---
SUM(bed_fact_table.license_beds) AS total_sicu_license_beds
FROM business_table
--- Inner join bed_fact_table with business_table on ims_org_id ---
JOIN bed_fact_table ON business_table.ims_org_id = bed_fact_table.ims_org_id
--- Only include rows where bed_id is 15 (SICU) ---
WHERE bed_fact_table.bed_id = 15
--- Group by hospital_name ---
GROUP BY business_table.business_name
--- Arrange in descending order ---
ORDER BY total_sicu_license_beds DESC
--- Only include top 10 hospitals ---
LIMIT 10;"

license_beds_SICU_df <- dbGetQuery(p2_connection, license_beds_SICU)
```

```{r, census_beds_separate}
# Sairah

census_beds_icu <- "SELECT business_name AS hospital_name, SUM(census_beds) AS total_icu_beds 
FROM combined_table
WHERE bed_id = 4
GROUP BY hospital_name
ORDER BY total_icu_beds DESC
LIMIT 10"

census_bed_icu <- dbGetQuery(p2_connection, census_beds_icu)

print(census_bed_icu)

ggplot(census_bed_icu, aes(x = hospital_name, y = total_icu_beds)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
  x = "Hospital Name",
  y = "Total Census ICU Beds",
  title = "Hospitals with the Highest Combined ICU Census Bed Counts",
  caption = "This bar chart looks at the hospitals with the higest ICU census bed counts."
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 9, color = "black"),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black"),
  plot.title = element_text(hjust = 0.1, face = "bold", margin = margin(b = 20)),
  plot.margin = margin(t = 30, r = 20, l = 20, b = 20)
  )

```

```{r}
census_beds_sicu <- "SELECT business_name AS hospital_name, SUM(census_beds) AS total_sicu_beds 
FROM combined_table
WHERE bed_id = 15
GROUP BY hospital_name
ORDER BY total_sicu_beds DESC
LIMIT 10"

census_bed_sicu <- dbGetQuery(p2_connection, census_beds_sicu)

print(census_bed_sicu)

ggplot(census_bed_sicu, aes(x = hospital_name, y = total_sicu_beds)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
  x = "Hospital Name",
  y = "Total Census SICU Beds",
  title = "Hospitals with the Highest Combined SICU Census Bed Counts",
  caption = "This bar chart looks at the hospitals with the higest SICU census bed counts."
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 9, color = "black"),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black"),
  plot.title = element_text(hjust = 0.1, face = "bold", margin = margin(b = 20)),
  plot.margin = margin(t = 30, r = 20, l = 20, b = 20)
  )

```

```{r, staffed_beds_separate}
get_staffed_beds_total <- "SELECT business_name AS hospital_name,
                                 --- Take the sum of the staffed beds in a new column total_icu_sicu_staffed_beds ---
                                 SUM(staffed_beds) AS total_icu_sicu_staffed_beds
                          FROM combined_table
                          --- Only include rows where bed_id is 4 (ICU) or 15 (SICU) ---
                          WHERE bed_id IN (4, 15)
                          --- Group by hospital_name ---
                          GROUP BY hospital_name
                          --- Arrange in descending order ---
                          ORDER BY total_icu_sicu_staffed_beds DESC
                          --- Only include top 10 hospitals ---
                          LIMIT 10;"

total_staffed_beds <- dbGetQuery(p2_connection, get_staffed_beds_total) %>%
  as_tibble() %>%
  mutate(
    total_icu_sicu_staffed_beds = as.numeric(total_icu_sicu_staffed_beds)
  )
total_staffed_beds
```

```{r staffed visual}

# Query for the visual
staffed_beds_separate_visual <- "SELECT business_name AS hospital_name,
                                 SUM(CASE WHEN bed_desc = 'ICU' THEN staffed_beds ELSE 0 END) AS ICU_beds,
                                 SUM(CASE WHEN bed_desc = 'SICU' THEN staffed_beds ELSE 0 END) AS SICU_beds,
                                 SUM(staffed_beds) AS total_icu_sicu_staffed_beds
                          FROM combined_table
                          --- Only include rows where bed_id is 4 (ICU) or 15 (SICU) ---
                          WHERE bed_desc IN ('ICU', 'SICU')
                          GROUP BY hospital_name
                          --- Arrange in descending order ---
                          ORDER BY total_icu_sicu_staffed_beds DESC;"

# Create new dataframe for visual
separate_staffed_beds_df <- dbGetQuery(p2_connection, staffed_beds_separate_visual) %>%
  as_tibble() %>%
  slice(1:10) %>%
  rename(ICU = ICU_beds,
         SICU = SICU_beds)
separate_staffed_beds_df

pivot_staffed_beds_df <- separate_staffed_beds_df %>%
  pivot_longer(cols = c(ICU, SICU), names_to = "bed_type", values_to = "bed_count")
```

``` {r}
# Stacked bar for better visualization of ICU and SICU beds for top 10 hospitals
ggplot(pivot_staffed_beds_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = bed_type)) +
  geom_bar(stat = "identity") +
  labs(
  x = "Hospital Name",
  y = "Staffed Bed Count",
  fill = "Bed",
  title = "Hospitals with the highest combined ICU and SICU staffed bed counts"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 9, color = "black"),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black"),
  plot.title = element_text(hjust = 0.1, face = "bold", margin = margin(b = 20)),
  plot.margin = margin(t = 30, r = 20, l = 20, b = 20)
  )
```

### Interpretation of Findings (Pt. 1)

### Follow-up Analysis for Leadership (identifying hospitals with both ICU and SICU volume sufficiency)

```{r, license_beds_combined}
# Thomas

```

```{r, census_beds_combined}
# Sairah

icu_sicu_census_beds <- "SELECT business_table.business_name AS hospital_name, SUM(bed_fact_table.census_beds) AS total_icu_sicu_census_beds
FROM business_table
JOIN bed_fact_table ON business_table.ims_org_id = bed_fact_table.ims_org_id 
WHERE bed_fact_table.bed_id IN (4,15)
GROUP BY business_table.business_name
ORDER BY total_icu_sicu_census_beds DESC
LIMIT 10;"

icu_sicu_census_beds <- dbGetQuery(p2_connection, icu_sicu_census_beds)

print(icu_sicu_census_beds)

library(ggplot2)

ggplot(icu_sicu_census_beds, aes(x = hospital_name, y = total_icu_sicu_census_beds)) + geom_bar(stat = "identity", fill = "light blue") + theme_minimal() + coord_flip() + labs(x = "Hospital Name", y = "Total ICU and SICU Census Beds") 

```

```{r, staffed_beds_combined}
staffed_beds_combined <- "SELECT business_name AS hospital_name,
                                 SUM(staffed_beds) AS total_icu_sicu_beds
                          FROM combined_table
                          --- Include ICU and SICU ---
                          WHERE (bed_id = 4 OR bed_id = 15)
                          GROUP BY hospital_name
                          --- Ensures at least 1 bed is counted in SICU and ICU ---
                          HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
                             AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
                          ORDER BY total_icu_sicu_beds DESC
                          LIMIT 10;"

total_staffed_beds <- dbGetQuery(p2_connection, staffed_beds_combined)
total_staffed_beds
```

```{r}
staffed_beds_combined_visual <- "SELECT business_name AS hospital_name,
                                 SUM(CASE WHEN bed_id = 4 THEN staffed_beds ELSE 0 END) AS ICU_beds,
                                 SUM(CASE WHEN bed_id = 15 THEN staffed_beds ELSE 0 END) AS SICU_beds,
                                 SUM(staffed_beds) AS total_icu_sicu_beds
                          FROM combined_table
                          --- Include ICU and SICU ---
                          WHERE (bed_id = 4 OR bed_id = 15)
                          GROUP BY hospital_name
                          --- Ensures at least 1 bed is counted in SICU and ICU ---
                          HAVING COUNT(DISTINCT CASE WHEN bed_id = 4 THEN 1 END) > 0
                             AND COUNT(DISTINCT CASE WHEN bed_id = 15 THEN 1 END) > 0
                          ORDER BY total_icu_sicu_beds DESC;"

staffed_beds_combined_visual_df <- dbGetQuery(p2_connection, staffed_beds_combined_visual) %>%
  as_tibble %>%
  slice(1:10) %>%
  rename(ICU = ICU_beds,
         SICU = SICU_beds)

staffed_beds_combined_visual_df

pivot_combined_staffed_beds_visual_df <- staffed_beds_combined_visual_df %>%
  pivot_longer(cols = c(ICU, SICU), names_to = "bed_type", values_to = "bed_count")
```

```{r}
# Stacked bar for better visualization of ICU and SICU beds for top 10 hospitals
ggplot(pivot_combined_staffed_beds_visual_df, aes(reorder(hospital_name, -bed_count), bed_count, fill = bed_type)) +
  geom_bar(stat = "identity") +
  labs(
  x = "Hospital Name",
  y = "Staffed Bed Count",
  fill = "Bed",
  title = "Hospitals with the highest counts of at least 1 ICU and SICU staffed beds"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  theme_classic() +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 9, color = "black"),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  legend.title = element_text(face = "bold"),
  axis.text.y = element_text(color = "black"),
  axis.title.y = element_text(color = "black"),
  plot.title = element_text(hjust = 0.1, face = "bold", margin = margin(b = 20)),
  plot.margin = margin(t = 30, r = 20, l = 20, b = 20)
  )
```

### Interpretation of Findings (Pt. 2)

TODO

### Final Recommendation

TODO
